Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-14 paper/sheaf_lics.tex
first draft of introduction

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{xspace}

@@

\newcommand{\IsType}[1]{\mathrm{Is\mymathhyphen}#1\mathrm{\mymathhyphen
    type}}
\newcommand{\modal}{\ensuremath{\ocircle}}

@@

  Sheafification is a popular tool in (higher) topos theory which
  allows to extend the internal logic of a topos with new
  principles. One of its most famous applications is the possibility
  to transform a topos into a boolean (classical) topos using the
  dense topology, which corresponds in essence to Gödel's double
  negation translation.
  % 
  The same construction has not been considered in Martin-Löf type
  theory because of a mismatch between topos theory and type theory. This
  mismatched has been fixed recently by considering homotopy type
  theory, an extension of Martin-Löf type theory with new
  principles inspired by category theory and homotopy theory, and
  which corresponds closely to higher topoi.
  % 
  In this paper, we give a computer-checked construction of
  Lawvere-Tierney sheafification in homotopy type theory, which allows
  in particular to give a meaning to the (propositional) law of
  excluded middle inside homotopy type theory, without axiom, being
  compatible with the full type-theoretic axiom of choice.

@@

Unfortunately, all those works do not share the generality and
simplicity of sheafification in topos theory.  
%
However, type theory is known to be quite closed to topos theory so
one could wonder why similar techniques have not been developed in the
field of type theory. 
%
The answer to this question has been given recently by the advent of
homotopy type theory~\cite{hottbook}, which is an extension of
Martin-Löf type theory with principles inspired by (higher) category
theory and homotopy theory, such as higher inductive
types~\cite{lumsdaine2011higher,lumsdaine2013higher} and
Voevodsky's univalence principle~\cite{kapulkin2012simplicial}.

This new point of view on type theory has revealed the
homotopy structure of types where for instance (homotopical)
propositions are just types with an irrelevant equality and
(homotopical) sets are types with a propositional equality. 
%
When restricted to such propositions and sets, type theory corresponds
quite closely to topos theory but the mismatch starts when considering
higher homotopical types. 
%
Fortunately, an higher version of topos theory has been developed
recently by mathematicians, synthesized in the monograph of Lurie on
higher topos theory~\cite{lurie}. 
%
Lurie has developed their all the tools that have been defined in
topos theory, but in an higher setting. In particular, the theory of
sheaves has been lifted to higher topos theory.
%
As the notion of higher topoi appears to correspond very closely
to homotopy type theory, this provides a new hope that tackling the problem of
extending the power of type theory using sheafification is actually possible.

@@

@@

The main contribution of this article is presents a computer check

\begin{itemize}
\item 
  modalities in HoTT
\item 
  Giraud's axiom
\end{itemize}

Plan of the paper~:
\begin{itemize}
\item In section~\ref{sec:hott}, we define the foundations on which
  our worked is based.
\item In section~\ref{sec:sheaves}, we explain the construction of the
  sheafification functor.
\item Finally, we discuss related works and present future works.
\end{itemize}
\section{Homotopy Type Theory}
\label{sec:hott}

@@

\nt{
  Il faut au minimum introduire proprement les n-Types, le squash
  sur les props 
}

@@

The last thing to prove is the cumulativity of our sheafifications~:
\begin{prop}
  If $T:\Type_n$, then $\modal T = \star \widehat T$ (where $\widehat
  T$ is $T$ seen as a $\Type_{n+1}$.
\end{prop}
It will imply that $\star$ is compatible with $\HProp$ in the sense of proposition~\ref{prop:hprop}.
\section{Future Works}
\label{sec:future-works}

@@

\section{Conclusion}
The conclusion goes here.

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{xspace}
\usepackage[colorlinks=true]{hyperref}

@@

\newcommand{\IsType}[1]{\mathrm{Is\mymathhyphen}#1\mathrm{\mymathhyphen
    type}}
\newcommand{\modal}{\ensuremath{\ocircle}}
\newcommand \True {\mathrm{true}}
\newcommand \False {\mathrm{false}}
\newcommand \closure[1] {\overline{#1}}
\newcommand \Char[1] {\chi_{#1}}%{\mathrm{char}(#1)}
\newcommand \E {\mathcal{E}}
\newcommand \Hom[1] {\mathrm{Hom}_{#1}}
\newcommand \Sh[1] {\mathrm{Sh}_{#1}}

@@

  Sheafification is a popular tool in topos theory which
  allows to extend the internal logic of a topos with new
  principles. One of its most famous applications is the possibility
  to transform a topos into a boolean topos using the
  dense topology, which corresponds in essence to Gödel's double
  negation translation.
  % 
  The same construction has not been considered in Martin-Löf type
  theory because of a mismatch between topos theory and type theory. This
  mismatched has been fixed recently by considering homotopy type
  theory, an extension of Martin-Löf type theory with new
  principles inspired by category theory and homotopy theory, and
  which corresponds closely to higher topoi.
  % 
  In this paper, we give a computer-checked construction of
  Lawvere-Tierney sheafification in homotopy type theory, which allows
  in particular to give a meaning to the (propositional) law of
  excluded middle inside homotopy type theory, without axiom, being
  compatible with the full type-theoretic axiom of choice.

@@

Unfortunately, all those works---although they provide very nice and
sophisticated theoretical results---do not share the generality and
simplicity of sheafification in topos theory and are not amenable to a
direct integration into proof assistant based on type theory such as
Coq or Agda.  
%
However, type theory is known to be quite closed to topos theory so
one could wonder why similar techniques have not been developed in the
field of type theory. 
%
The answer to this question has been given recently by the advent of
homotopy type theory~\cite{hottbook}, which is an extension of
Martin-Löf type theory with principles inspired by (higher) category
theory and homotopy theory, such as higher inductive
types~\cite{lumsdaine2011higher,lumsdaine2013higher} and
Voevodsky's univalence principle~\cite{kapulkin2012simplicial}. 
%

This new point of view on type theory has revealed the
homotopy structure of types where for instance homotopical
propositions (or $\Type_{-1}$) are just types with an irrelevant equality and
homotopical sets (or $\Type_{0}$)  are types with a propositional
equality and so on for $\Type_{n}$.
%
The development of univalence has also shed some light on the
difficulty to make the AC and EM coexist in type theory. Indeed, it
has been shown that a naive (non-propositional) version of EM is
inconsistent with univalence.

When restricted to homotopical propositions and sets, homotopy type
theory corresponds quite closely to topos theory but the mismatch
starts when considering higher homotopical types.
%
Fortunately, a higher version of topos theory has been developed
recently by mathematicians, synthesized in the monograph of Lurie on
higher topos theory~\cite{lurie}. 
%
Lurie has developed all the tools that have been defined in
topos theory, but in a higher setting. In particular, the theory of
sheaves has been lifted to higher topos theory.
%
As the notion of higher topoi appears to correspond very closely
to homotopy type theory, this provides a new hope that tackling the problem of
extending the power of type theory using sheafification is actually possible.

@@

Lawvere-Tierney sheafification in a topos $\E$ is based on an abstract
point of view on the topology to be considered, being simply an
endomorphism on the classifying object $\Omega$ of $\E$  
%
$$
j : \Omega \rightarrow \Omega
$$
%
that is required to preserve $\True$ ($j \ \True = \True$), to be
idempotent ($j \circ j = j$) and compatible with products ($j \circ
\wedge = \wedge \circ (j, j)$).
%
A typical example is given by double negation.

Every Lawvere-Tierney topology $j$ induces a closure operator
$A \mapsto \closure{A}$ on subojects. If we see a subobject $A$ of $E$
as a characteristic function $\Char{A}$, the closure $\closure{A}$
corresponds to the subobject of $E$ whose characteristic function is 
%
$$
\Char{\closure{A}} = j \circ \Char{A}.
$$
%
A suboject $A$ of $E$ is said to
be dense when $\closure{A} = E$.

The idea is then to define sheaves in $\E$ as objects of $\E$ for
which it is impossible to make a distinction between objects and their
dense subobjects. This idea is formalized by saying that for every
dense subobject $A$ of $E$, the following canonical map is an
isomorphism
%
\begin{equation}\label{equ:sheaf_def}
\Hom{\E}(E,F) \rightarrow \Hom{\E}(A,F).
\end{equation}
%
On can show that $\Sh{\E}$, the full sub-category of $\E$ given by
sheaves, is again a topos, with classifying object
%
$$
\Omega_j = \{ P \in \E \ | \ j P  \leftrightarrow P \}.
$$
%
Thus, in case of the double negation, the resulting topos is boolean  
and admits classical reasoning.

But furthermore, one can defined a left adjoint, the sheafification
functor, 
%
$$
a_j : \E \rightarrow \Sh{\E}
$$
to the inclusion functor, which exhibits $\Sh{\E}$ as a reflective
subcatefory of $\E$ (which is a particular case of localization). This
means that logical principles valid in $\E$ are still valid in
$\Sh{\E}$.

@@

The main contribution of this article is the development of a
computer-checked definition of Lawvere-Tierney sheafification in the
setting of homotopy type theory.
%
This provides a way to give a meaning to the propositional law of
excluded middle, and more generally, combined with the previous work on
forcing in type theory~\cite{jaber2012extending}, it allows to reuse
results derived for topos theory, such as independence results.

To extend Lawvere-Tierney sheafification to homotopy type theory, the
first thing to realize is that the construction can not be done in one
step anymore. It must be done by induction on the level of homotopy of
types. More precisely, the first layer of sheafification is defined
for $\Type_{0}$ , given a topology on $\Type_{-1}$. This part
corresponds to sheafification for topoi. Then, assuming that the
sheafification has been constructed up to level $n$, one can defined
the sheafification for $\Type_{n+1}$.
%
This inductive process in the definition of sheafification is similar
to construction in the Grothendieck setting, where the
sheafification is obtained by applying the plus-construction $n+1$ to
$\Type_{n}$~\cite[Chapter~6]{lurie}.

This inductive step requires to formalize the notion of left-exact
reflective subuniverse, which corresponds to a stratified
version of left-exact modalities introduced
in~\cite[Chapter~7]{hottbook}. A Lawvere-Tierney topology can thus be
seen as a left-exact modality on $\Type_{-1}$ and sheafification as
a way to extend it as a left-exact modality on $\Type_{n}$ for any $n$.

Compare to the definition of sheafification in topos theory, the main
change comes from the fact that in homotopy type theory, there is not
a single object classifier but rather an $n$-object classifier for
each $n\geq -2$ which classifies functions with $n$-truncated homotopy
fibers~\cite{sets_in_hott}. This means that when monomorphisms are used in topos
theoretic definition, the correct generalization may be $n$-truncated homotopy
fibers or monomorphims (\ie $-1$-truncated homotopy
fibers) depending on the context.

For instance, the notion of separated objects (which is a technical
tool in the construction of the sheafification), for which the morphism
(\ref{equ:sheaf_def}) is just a monomorphism, is generalized in two
different ways.  
%
The subobjects to be considered are given by $n$-truncated fibers,
while the morphism (\ref{equ:sheaf_def}) is still required to be a
monomorphism.

In other technical difficulty is the fact that sheafification in topos
theory uses the fact that epimorphisms are coequalizers of their
kernel pairs. This properties does not holds in higher topoi, but it
can be replaced by Giraud's axiom which basically allows to replace 
 kernel pairs by \v{C}ech nerves.

Finally, one of the main difficulty in our formalization is that many
equlity which are trival (b yproof irrelevance) in the topos theoretic
setting becomes relevant and requires a carreful use of univalence
with appropriate isomorphisms.

\subsection{Plan of the Paper}
\begin{itemize}
\item Section~\ref{sec:hott}, we define the foundations on which
  our worked is based.
\item Section~\ref{sec:sheaves}, we explain the construction of the
  sheafification functor.
\item Section~\ref{sec:formalization} presents the details of the
  formalization in Coq.
\item Finally, we present future works.
\end{itemize}

\section{Preliminaries on Homotopy Type Theory}
\label{sec:hott}

@@

@@

The last thing to prove is the cumulativity of our sheafifications~:
\begin{prop}
  If $T:\Type_n$, then $\modal T = \star \widehat T$ (where $\widehat
  T$ is $T$ seen as a $\Type_{n+1}$.
\end{prop}
It will imply that $\star$ is compatible with $\HProp$ in the sense of proposition~\ref{prop:hprop}.

@@

\section{Formalization} 
\label{sec:formalization}

A Coq formalization is available at
\url{https://github.com/KevinQuirin/sheafification/}.

\nt{describe how many line for each, be clear about what has not been formalized}

\section{Conclusion and Future Works}
\label{sec:future-works}
